---
title: "Diversity Analysis with PhyloIgSeq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diversity Analysis with PhyloIgSeq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

This vignette covers PhyloIgSeq's `phyloseq`-extension side: filtering, alpha/beta
diversity, and compositional heatmaps, on a standard 16S `phyloseq` object. For the
package's namesake Ig-coating scoring pipeline, see
`vignette("igseq-scoring", package = "PhyloIgSeq")` instead.

```{r setup}
library(PhyloIgSeq)
library(phyloseq)
```

## Data

The examples below use `ps_16s_refinement`, a bundled `phyloseq` object comparing
different 16S sequencing protocols: 515 taxa across 37 samples, with a taxonomy table,
sample metadata, and a phylogenetic tree.

```{r data}
data(ps_16s_refinement)
ps_16s_refinement
sample_variables(ps_16s_refinement)
```

## Filtering

`filter_reads()` removes low-depth samples and low-abundance taxa by total read count.
`plot_reads()` shows the read-count distributions a threshold would act on, so you can
pick one before filtering:

```{r plot-reads}
plot_reads(ps_16s_refinement, min_sample_sum = 1000, min_taxa_sum = 5)
```

```{r filter}
ps <- filter_reads(ps_16s_refinement, min_sample_sum = 1000, min_taxa_sum = 5)
ps
```

The rest of this vignette uses `ps`, the filtered object.

## Alpha diversity

`get_alpha_diversity()` computes one or more per-sample diversity measures;
`plot_alpha_diversity()` renders the result, optionally grouped/colored by a
`sample_data` column:

```{r alpha}
alpha_div <- get_alpha_diversity(ps, measure = "Shannon")
head(alpha_div$diversity)
plot_alpha_diversity(alpha_div, x = "Concentration", group = "SampleOrigin")
```

`alpha_diversity()` is a thin wrapper chaining both calls, for when you don't need the
intermediate `get_alpha_diversity()` result:

```{r alpha-wrapper}
alpha_diversity(ps, x = "Concentration", group = "SampleOrigin")
```

## Beta diversity

### Unconstrained ordination

`get_beta_diversity()` ordinates samples by inter-sample dissimilarity — distance-based
(PCoA, tSNE, UMAP), direct-on-abundance (NMDS, PCA, CA, DCA), or constrained (CCA, RDA,
dbRDA; see below). `plot_beta_diversity()` renders the result, `scree_plot()` shows how
much variance each axis explains, and `stat_beta_diversity()` runs a PERMANOVA against a
grouping variable:

```{r beta-unconstrained}
beta_div <- get_beta_diversity(ps, method = "pcoa", dist = "bray")
plot_beta_diversity(beta_div, label_name = "SampleOrigin", ellipses = TRUE)
scree_plot(beta_div$eigen_values)
stat_beta_diversity(beta_div, label_name = "SampleOrigin")$p_value
```

`beta_diversity()` chains all three in one call:

```{r beta-wrapper}
beta_diversity(ps, group = "SampleOrigin", stat = TRUE)
```

### Constrained ordination

Passing `method = "CCA"`/`"RDA"`/`"dbRDA"` with a `model` formula constrains the
ordination to a `sample_data` variable, e.g. to see how much of the community
composition a technical covariate like the mechanical lysis protocol explains:

```{r beta-constrained}
beta_cca <- get_beta_diversity(
  ps,
  taxrank = "Genus",
  method = "CCA",
  model = "Mechanical_Lyse"
)
plot_beta_diversity(beta_cca, label_name = "Mechanical_Lyse")
```

## Compositional heatmap

`get_phylo_heatmap()` prepares an abundance matrix, sample dendrogram, and taxon
ordering; `plot_phylo_heatmap()` renders it as a `ComplexHeatmap::Heatmap()`, limited to
the top `nb_top_taxa` by abundance:

```{r heatmap, fig.height = 6}
heatmap_data <- get_phylo_heatmap(ps, taxrank_for_heatmap = "Genus")
plot_phylo_heatmap(heatmap_data, top_annotation_vars = "Protocol", nb_top_taxa = 20)
```

## Working with large datasets: the sparse engine

Everything above also works, unchanged, on a *sparse* `phyloseq` object.
`as_sparse_phyloseq()` replaces a dense OTU table with a `dgCMatrix`-backed one;
`sparse_distance()`/`sparse_unifrac()` then compute Bray-Curtis, Jaccard, (weighted)
UniFrac and other distances directly on it, without ever densifying the table. They are
drop-in, numerically equivalent replacements for `phyloseq::distance()`/
`phyloseq::UniFrac()`:

```{r sparse}
ps_sparse <- as_sparse_phyloseq(ps)
class(otu_table(ps_sparse))

bc_sparse <- sparse_distance(ps_sparse, method = "bray")
```

UniFrac needs a `phy_tree`. When no sequencing-based tree is available,
`get_taxonomy_tree()` builds an approximate one from `tax_table` ranks alone — a
heuristic, not a substitute for a real phylogeny, but enough to keep UniFrac-family
distances available:

```{r taxonomy-tree}
phy_tree(ps_sparse) <- get_taxonomy_tree(ps_sparse)
sparse_unifrac(ps_sparse, method = "wunifrac")
```

`get_beta_diversity()`'s distance-based methods (PCoA/tSNE/UMAP) and
`get_phylo_heatmap()` already call `sparse_distance()` internally, so no other code
needs to change to benefit — the recommended workflow for a large ASV/OTU table is
simply to convert it with `as_sparse_phyloseq()` once, right after filtering, and pass
the sparse object into every function downstream:

```{r sparse-workflow}
beta_sparse <- get_beta_diversity(ps_sparse, method = "PCoA", dist = "bray")
all.equal(beta_sparse$coords, beta_div$coords)
```

## Session info

```{r session}
sessionInfo()
```
