---
title: "Ig-Coating Scoring with PhyloIgSeq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ig-Coating Scoring with PhyloIgSeq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

This vignette covers the package's namesake pipeline: quantifying immunoglobulin (Ig)
coating per taxon per sample from IgSeq sort-fraction 16S data (Ig+/Ig-, optionally a
second Ig- replicate and/or a pre-sort fraction). For general `phyloseq` diversity
analysis, see `vignette("diversity-analysis", package = "PhyloIgSeq")` instead — the two
are complementary and, as the last section here shows, IgSeq scores can themselves be
fed back into that vignette's diversity functions.

```{r setup}
library(PhyloIgSeq)
library(phyloseq)
```

## Data

The examples below use `ps_igseq`, a small, anonymized toy `phyloseq` object from a
mouse IgA-Seq experiment: 4703 taxa, 38 samples (rows), each belonging to one of 8
biological samples split across up to four sort fractions (`pos`, `neg1`, `neg2`,
`whole`, recorded in the `sorting_fraction` column). `whole` (pre-sort) and the
`ig_pheno` sample-metadata column (Ig+ frequency phenotype) are synthetic additions
that let this toy dataset also demonstrate the positive-fraction-only workflow (see
below) — like `sex`/`age`, they carry no biological meaning.

```{r data}
data(ps_igseq)
ps_igseq
sample_data(ps_igseq)
```

## Building a PhyloIgSeq object

`getPhyloIgSeq()` is the pipeline's entry point. It reshapes the per-fraction rows into
one profile per biological sample (`sample_id_name`), rarefies fractions to a comparable
depth, and computes every score named in `scores` for each taxon:

```{r build}
pis <- getPhyloIgSeq(
  physeq = ps_igseq,
  sample_id_name = "sample_id",
  fraction_id_name = "sorting_fraction",
  positive_fraction_name = "pos",
  first_negative_fraction_name = "neg1",
  second_negative_fraction_name = "neg2",
  confidence_levels = c(0.95, 0.99),
  scores = c("slide_z", "palm", "kau")
)
pis
```

Two samples in this toy dataset (`sample_6`, `sample_7`) have duplicated fraction rows
and are excluded automatically, with a `warning()` naming which sample and why —
`getPhyloIgSeq()` skips rows it can't interpret rather than aborting the whole batch, so
watch for this warning on your own data too.

The result is a `PhyloIgSeq` object: `ig_coating` holds one row per taxon/sample, with
the requested scores as columns (also recorded in `score_names`) alongside supporting
fraction/diagnostic columns. `get_ig_score()` pulls out a single score without having to
know which of the other columns are metadata:

```{r get-score}
head(get_ig_score(pis, score_name = "slide_z", sample_ids = "sample_1"))
```

### Positive-fraction-only mode

Everything above compares Ig+ against a negative fraction. If you only sequenced a
positive fraction, `prob_index` (see below) can still be computed from just the positive
fraction, a pre-sort fraction, and an Ig+ frequency phenotype — pass
`first_negative_fraction_name = NULL` and supply `presorting_fraction_name`/
`ig_freq_name` instead. Every other score (and `slide_z`'s MA-plot geometry) is
fundamentally a positive-vs-negative comparison, so `getPhyloIgSeq()` drops `slide_z`
from `scores` (with a `warning()`) if it's requested without a negative fraction, and
errors if neither a negative fraction nor both `presorting_fraction_name`/`ig_freq_name`
are supplied:

```{r pos-only}
pis_pos_only <- getPhyloIgSeq(
  physeq = ps_igseq,
  sample_id_name = "sample_id",
  fraction_id_name = "sorting_fraction",
  positive_fraction_name = "pos",
  presorting_fraction_name = "whole",
  ig_freq_name = "ig_pheno",
  scores = "prob_index"
)
pis_pos_only
head(get_ig_score(pis_pos_only, score_name = "prob_index", sample_ids = "sample_1"))
```

### Per-fraction Ig+ frequencies

`ig_pheno` above is one value per biological sample: an overall P(Ig+), repeated on each
of that sample's fraction rows. That's the default, `ig_freq_layout = "wide"`, and it is
all `prob_index`/`prob_ratio` need.

Flow-cytometry phenotyping is often done on each sorted fraction separately, though —
giving not just the pre-sort P(Ig+) but also how pure the Ig+ fraction actually came out
and how much Ig+ material was left behind in the Ig− one. Recording that means one
`sample_data` column whose value *differs between the fractions of a sample*, which
`ig_freq_layout = "long"` reads accordingly:

```{r long-layout}
ps_long <- ps_igseq
sd <- as(sample_data(ps_long), "data.frame")
# Ig+ frequency measured in each fraction rather than once per sample. neg1/neg2
# are two aliquots of the same negative fraction, so they share a measurement.
sd$ig_pheno_fraction <- c(
  whole = 0.40, pos = 0.80, neg1 = 0.10, neg2 = 0.10
)[sd$sorting_fraction]
sample_data(ps_long) <- sample_data(sd)

pis_long <- getPhyloIgSeq(
  physeq = ps_long,
  sample_ids = c("sample_1", "sample_2", "sample_3"),
  sample_id_name = "sample_id",
  fraction_id_name = "sorting_fraction",
  positive_fraction_name = "pos",
  first_negative_fraction_name = "neg1",
  second_negative_fraction_name = "neg2",
  presorting_fraction_name = "whole",
  ig_freq_name = "ig_pheno_fraction",
  ig_freq_layout = "long",
  scores = c("prob_index", "purity_corrected_prob_index")
)
pis_long@sample_data[, c(
  "sample_id", "presort_ig_freq", "pos_ig_freq", "neg1_ig_freq", "neg2_ig_freq"
)]
```

Whichever layout is used, the resolved (unit-converted, range-checked) values are stored
per sample in `sample_data`, so they can be inspected, exported, and used to color or
facet downstream plots. Under `"wide"` only `presort_ig_freq` is populated — a single
overall P(Ig+) says nothing about how pure either sorted fraction turned out to be.

The two `purity_corrected_*` scores need those in-fraction measurements, so they are only
computable under `"long"` (with both a negative and a pre-sort fraction); requesting them
otherwise drops them from `scores` with a `warning()`. They remain **experimental** — see
`?compute_ig_score` and the comments in its source, which spell out the sorting parameter
they depend on and the caveats of deriving it.

Picking `"wide"` for a column that is in fact per-fraction is the one trap here: the value
can't be collapsed to a single per-sample number, so it goes unused. `getPhyloIgSeq()`
warns per sample and points at `"long"` rather than silently returning `NA` scores.

### Multiple positive fractions

If several positive fractions were sorted from the same samples against one shared
negative/pre-sort fraction — e.g. multiple phage-display sort rounds or output pools —
`positive_fraction_name` can be a character vector instead of a single value. Each
(sample, positive fraction) combination becomes its own row. Crucially, the negative
fraction is rarefied *once*, jointly across every positive fraction of a sample, rather
than independently for each one — so different positive fractions of the same sample are
compared against their shared negative fraction at one common depth.

`ps_phage_display` is a toy dataset built for exactly this scenario: 8 panning campaigns,
each with one shared `"input"` (naive/unselected) library and three sequential positive
output rounds (`"round1"`/`"round2"`/`"round3"`), progressively enriched for a handful of
campaign-specific clones:

```{r multi-pos}
data(ps_phage_display)
pis_multi <- getPhyloIgSeq(
  physeq = ps_phage_display,
  sample_id_name = "sample_id",
  fraction_id_name = "phage_fraction",
  positive_fraction_name = c("round1", "round2", "round3"),
  first_negative_fraction_name = "input",
  scores = c("slide_z", "palm", "kau")
)
pis_multi
head(pis_multi@sample_data)
```

Rows are keyed by a synthetic `sample_id` (`"<sample_id>_<positive fraction>"`), and
`sample_data` gains `original_sample_id`/`positive_fraction_name` columns recording where
each row came from — usable directly as a facet/color variable in downstream diversity
analyses (`vignette("diversity-analysis", package = "PhyloIgSeq")`). The raw
positive-fraction abundance column is named `positive_fraction_abundance` (rather than
after any one fraction's own value) so it stays populated for every row regardless of
which positive fraction it came from.

Rows from the same original sample across different positive fractions are **not**
independent replicates: they share the same underlying biology and the same negative
fraction. With a single `positive_fraction_name` (the default, and everything else in
this vignette), output is unchanged: no suffix, no extra columns.

## Classical Ig scores

`compute_ig_score()` is the lower-level function behind `palm`/`kau`/`prob_index`/
`prob_ratio` (the names in `IG_SCORES`): closed-form indices computed directly from
pos/neg(/pre) abundance vectors, with no sliding window. Useful on its own when you
already have abundance vectors and don't need the full `PhyloIgSeq` pipeline:

```{r classical-scores}
IG_SCORES

pos <- c(120, 45, 0, 300)
neg <- c(80, 60, 10, 250)
compute_ig_score(method = "kau", pos = pos, neg = neg)
compute_ig_score(method = "palm", pos = pos, neg = neg)
```

## The sliding Z-score

`slide_z` is this package's own scoring statistic. Classical scores like Palm/Kau assume
one global null distribution for "not coated"; the sliding Z-score instead estimates a
*local* null along the abundance axis (MA-plot-style), so that low- and high-abundance
taxa are judged against a null appropriate to their own abundance range rather than a
single package-wide one. `plot_slide_z()` shows this directly, plotting each taxon's
observed vs. null log-ratio with confidence ellipses (from `confidence_levels`, set
above) marking the local null at each abundance level:

```{r slide-z}
plot_slide_z(pis, sample_ids = "sample_1")
```

## Visualizing and agglomerating scores

`plot_ig_score()` renders a score agglomerated by taxon and sample group, as a
boxplot/violin or a bubble plot:

```{r plot-ig-score}
PhyloIgSeq::plot_ig_score(pis, plot_type = "boxplot", score_name = "slide_z", taxrank_score = "Genus")
```

`agglomPhyloIgSeq()` performs that agglomeration on the `PhyloIgSeq` object itself
(every score in `ig_coating`, not just the one being plotted), optionally weighting or
filtering by a fraction-abundance column:

```{r agglom}
pis_genus <- agglomPhyloIgSeq(
  pis,
  abundance_fraction = "pos",
  taxrank = "Genus",
  agglom_method = "median"
)
head(pis_genus@ig_coating)
```

## Exporting scores for downstream analysis

`PhyloIgSeq_to_phyloseq()` reshapes one score column into a samples-by-taxa "OTU table"
of score values and wraps it back into a standard `phyloseq` object, so the rest of the
`phyloseq` ecosystem (and PhyloIgSeq's own diversity functions, see below) can be applied
to Ig-coating scores directly. It offers two independent code paths, chosen with
`imputation_method`:

- For a small or already-agglomerated score matrix (e.g. `pis_genus` above), a dense
  matrix with ordinary imputation (`"KNN"`, `"Central Tendency"`, or, simplest, replacing
  missing scores with `0`) is usually fine:

  ```{r export-dense}
  ps_scores_dense <- PhyloIgSeq_to_phyloseq(
    pis_genus,
    score_name = "slide_z",
    imputation_method = "Replace NA with 0"
  )
  ps_scores_dense
  ```

- A **raw ASV/OTU-level** score matrix, by contrast, is typically mostly missing (most
  taxa aren't observed with a usable abundance in most samples) — densifying it first
  would be wasteful and would let arbitrary `0`s dominate downstream distances.
  `imputation_method = "SVD"` avoids this: missing entries stay structurally missing in
  an `incomplete_otu_table` (see `?incomplete_otu_table-class`), backed by
  `softImpute::Incomplete()`, and a low-rank approximation
  (`softImpute::softImpute()`, tuned by `svd_rank`/`svd_lambda`) is used to impute values
  only on demand. This is the recommended path for large, sparse score matrices:

  ```{r export-svd}
  ps_scores_svd <- PhyloIgSeq_to_phyloseq(
    pis,
    score_name = "slide_z",
    imputation_method = "SVD",
    svd_rank = 5L
  )
  class(otu_table(ps_scores_svd))
  ```

## IgSeq diversity: feeding scores back into the diversity functions

Because `PhyloIgSeq_to_phyloseq()`'s output is a standard `phyloseq` object, it plugs
directly into the functions from `vignette("diversity-analysis")` — with
`from_igseq = TRUE`, `get_alpha_diversity()` instead computes "richness" by bucketing
taxa into down-/not-significantly-/up-coated groups by their score, an Ig-coating
analogue of alpha diversity:

```{r igseq-alpha}
igseq_alpha <- get_alpha_diversity(ps_scores_dense, from_igseq = TRUE)
head(igseq_alpha$diversity)
plot_igseq_richness(igseq_alpha, group = "sex", color = "sex")
```

`get_beta_diversity()`/`plot_beta_diversity()` need no `from_igseq` switch at all — an
Ig-score "OTU table" ordinates exactly like an abundance one, giving an IgSeq beta
diversity view of which samples have similar coating profiles (here using the SVD-backed
export, so the ordination never densifies the underlying score matrix):

```{r igseq-beta}
igseq_beta <- get_beta_diversity(ps_scores_svd, method = "PCoA", dist = "euclidean")
plot_beta_diversity(igseq_beta, label_name = "sex")
```

## Session info

```{r session}
sessionInfo()
```
