---
title: "Ig-Coating Scoring with PhyloIgSeq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ig-Coating Scoring with PhyloIgSeq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

This vignette covers the package's namesake pipeline: quantifying immunoglobulin (Ig)
coating per taxon per sample from IgSeq sort-fraction 16S data (Ig+/Ig-, optionally a
second Ig- replicate and/or a pre-sort fraction). For general `phyloseq` diversity
analysis, see `vignette("diversity-analysis", package = "PhyloIgSeq")` instead — the two
are complementary and, as the last section here shows, IgSeq scores can themselves be
fed back into that vignette's diversity functions.

```{r setup}
library(PhyloIgSeq)
library(phyloseq)
```

## Data

The examples below use `ps_igseq`, a small, anonymized toy `phyloseq` object from a
mouse IgA-Seq experiment: 4703 taxa, 30 samples (rows), each belonging to one of 8
biological samples split across up to three sort fractions (`Pos`, `Neg1`, `Neg2`,
recorded in the `sorting_fraction` column).

```{r data}
data(ps_igseq)
ps_igseq
sample_data(ps_igseq)
```

## Building a PhyloIgSeq object

`getPhyloIgSeq()` is the pipeline's entry point. It reshapes the per-fraction rows into
one profile per biological sample (`sample_id_name`), rarefies fractions to a comparable
depth, and computes every score named in `scores` for each taxon:

```{r build}
pis <- getPhyloIgSeq(
  physeq = ps_igseq,
  sample_id_name = "sample_id",
  fraction_id_name = "sorting_fraction",
  positive_fraction_name = "Pos",
  first_negative_fraction_name = "Neg1",
  second_negative_fraction_name = "Neg2",
  confidence_levels = c(0.95, 0.99),
  scores = c("slide_z", "palm", "kau")
)
pis
```

Two samples in this toy dataset (`sample_6`, `sample_7`) have duplicated fraction rows
and are excluded automatically, with a `warning()` naming which sample and why —
`getPhyloIgSeq()` skips rows it can't interpret rather than aborting the whole batch, so
watch for this warning on your own data too.

The result is a `PhyloIgSeq` object: `ig_coating` holds one row per taxon/sample, with
the requested scores as columns (also recorded in `score_names`) alongside supporting
fraction/diagnostic columns. `get_ig_score()` pulls out a single score without having to
know which of the other columns are metadata:

```{r get-score}
head(get_ig_score(pis, score_name = "slide_z", sample_ids = "sample_1"))
```

## Classical Ig scores

`compute_ig_score()` is the lower-level function behind `palm`/`kau`/`prob_index`/
`prob_ratio` (the names in `IG_SCORES`): closed-form indices computed directly from
pos/neg(/pre) abundance vectors, with no sliding window. Useful on its own when you
already have abundance vectors and don't need the full `PhyloIgSeq` pipeline:

```{r classical-scores}
IG_SCORES

pos <- c(120, 45, 0, 300)
neg <- c(80, 60, 10, 250)
compute_ig_score(method = "kau", pos = pos, neg = neg)
compute_ig_score(method = "palm", pos = pos, neg = neg)
```

## The sliding Z-score

`slide_z` is this package's own scoring statistic. Classical scores like Palm/Kau assume
one global null distribution for "not coated"; the sliding Z-score instead estimates a
*local* null along the abundance axis (MA-plot-style), so that low- and high-abundance
taxa are judged against a null appropriate to their own abundance range rather than a
single package-wide one. `plot_slide_z()` shows this directly, plotting each taxon's
observed vs. null log-ratio with confidence ellipses (from `confidence_levels`, set
above) marking the local null at each abundance level:

```{r slide-z}
plot_slide_z(pis, sample_ids = "sample_1")
```

## Visualizing and agglomerating scores

`plot_ig_score()` renders a score agglomerated by taxon and sample group, as a
boxplot/violin or a bubble plot:

```{r plot-ig-score}
PhyloIgSeq::plot_ig_score(pis, plot_type = "boxplot", score_name = "slide_z", taxrank_score = "Genus")
```

`agglomPhyloIgSeq()` performs that agglomeration on the `PhyloIgSeq` object itself
(every score in `ig_coating`, not just the one being plotted), optionally weighting or
filtering by a fraction-abundance column:

```{r agglom}
pis_genus <- agglomPhyloIgSeq(
  pis,
  abundance_fraction = "Pos",
  taxrank = "Genus",
  agglom_method = "median"
)
head(pis_genus@ig_coating)
```

## Exporting scores for downstream analysis

`PhyloIgSeq_to_phyloseq()` reshapes one score column into a samples-by-taxa "OTU table"
of score values and wraps it back into a standard `phyloseq` object, so the rest of the
`phyloseq` ecosystem (and PhyloIgSeq's own diversity functions, see below) can be applied
to Ig-coating scores directly. It offers two independent code paths, chosen with
`imputation_method`:

- For a small or already-agglomerated score matrix (e.g. `pis_genus` above), a dense
  matrix with ordinary imputation (`"KNN"`, `"Central Tendency"`, or, simplest, replacing
  missing scores with `0`) is usually fine:

  ```{r export-dense}
  ps_scores_dense <- PhyloIgSeq_to_phyloseq(
    pis_genus,
    score_name = "slide_z",
    imputation_method = "Replace NA with 0"
  )
  ps_scores_dense
  ```

- A **raw ASV/OTU-level** score matrix, by contrast, is typically mostly missing (most
  taxa aren't observed with a usable abundance in most samples) — densifying it first
  would be wasteful and would let arbitrary `0`s dominate downstream distances.
  `imputation_method = "SVD"` avoids this: missing entries stay structurally missing in
  an `incomplete_otu_table` (see `?incomplete_otu_table-class`), backed by
  `softImpute::Incomplete()`, and a low-rank approximation
  (`softImpute::softImpute()`, tuned by `svd_rank`/`svd_lambda`) is used to impute values
  only on demand. This is the recommended path for large, sparse score matrices:

  ```{r export-svd}
  ps_scores_svd <- PhyloIgSeq_to_phyloseq(
    pis,
    score_name = "slide_z",
    imputation_method = "SVD",
    svd_rank = 5L
  )
  class(otu_table(ps_scores_svd))
  ```

## IgSeq diversity: feeding scores back into the diversity functions

Because `PhyloIgSeq_to_phyloseq()`'s output is a standard `phyloseq` object, it plugs
directly into the functions from `vignette("diversity-analysis")` — with
`from_igseq = TRUE`, `get_alpha_diversity()` instead computes "richness" by bucketing
taxa into down-/not-significantly-/up-coated groups by their score, an Ig-coating
analogue of alpha diversity:

```{r igseq-alpha}
igseq_alpha <- get_alpha_diversity(ps_scores_dense, from_igseq = TRUE)
head(igseq_alpha$diversity)
plot_igseq_richness(igseq_alpha, group = "sex", color = "sex")
```

`get_beta_diversity()`/`plot_beta_diversity()` need no `from_igseq` switch at all — an
Ig-score "OTU table" ordinates exactly like an abundance one, giving an IgSeq beta
diversity view of which samples have similar coating profiles (here using the SVD-backed
export, so the ordination never densifies the underlying score matrix):

```{r igseq-beta}
igseq_beta <- get_beta_diversity(ps_scores_svd, method = "PCoA", dist = "euclidean")
plot_beta_diversity(igseq_beta, label_name = "sex")
```

## Session info

```{r session}
sessionInfo()
```
