---
title: "Introduction to PhyloIgSeq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to PhyloIgSeq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(PhyloIgSeq)
```

## Overview

PhyloIgSeq provides a unified workflow for microbiota diversity analysis and
immunoglobulin (Ig) coating quantification from 16S amplicon sequencing data.
The package takes a `phyloseq` object — typically the output of a DADA2
pipeline — and guides the user through filtering, alpha/beta diversity
analysis, compositional visualisation, and Ig coating scoring.

This vignette walks through a complete end-to-end analysis using the bundled
example datasets.

---

## Data

```{r data}
# TODO: load example dataset(s)
# data("...")
```

---

## Filtering

Apply sample- and taxa-level filters before downstream analysis.

```{r filter}
# TODO: show filter_samples() / filter_taxa() usage on ps_16s_human
```

---

## Alpha diversity

Alpha diversity summarises within-sample richness and evenness.

```{r alpha}
# TODO: plot_alpha_diversity(ps_16s_human, measures = c("Observed", "Shannon"), group_var = "...")
```

---

## Beta diversity

Beta diversity compares community composition between samples via ordination.

```{r beta-unconstrained}
# TODO: unconstrained ordination (PCoA / PCA / MDS) on ps_16s_human
```

```{r beta-constrained}
# TODO: constrained ordination (RDA / CCA) with environmental variables
```

---

## Compositional analysis

### Barplots

```{r barplot}
# TODO: taxonomic barplot at phylum or genus level
```

### Heatmaps

```{r heatmap}
# TODO: heatmap of top N taxa across samples
```

---

## Ig coating analysis

The `PhyloIgSeq` S4 class stores Ig-Seq fractions and coating scores together
with optional taxonomic and sample metadata.

### Building a PhyloIgSeq object

```{r igseq-build}
# TODO: construct a PhyloIgSeq object from ps_mice_iga
# igseq <- new("PhyloIgSeq", ...)
```

### Computing Ig scores

`compute_ig_score()` supports several established scoring methods:

```{r ig-scores}
# TODO: compute Palm and Kau scores; compare across methods
# scores <- compute_ig_score(method = "kau", pos = ..., neg = ...)
```

### Sliding Z-score

The sliding Z-score is a novel per-taxon coating metric that accounts for
variation in Ig+ / Ig− frequency across the abundance range.

```{r slide-z}
# TODO: compute and visualise sliding Z-score
```

---

## Export

Results can be exported to Excel-ready tables for use in downstream modelling
(e.g., the Feature Selector App).

```{r export, eval = FALSE}
# TODO: export metadata, taxonomy, and abundance tables
```

---

## Session info

```{r session}
sessionInfo()
```
